\documentclass[10pt,twocolumn,letterpaper]{article}
\usepackage[pagenumbers]{cvpr}
\usepackage{times}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{xcolor}
\usepackage{bm}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{tikz}
\usetikzlibrary{positioning,arrows.meta,fit,calc,shapes.geometric,decorations.pathreplacing}
\usepackage{subcaption}
\usepackage[pagebackref=true,breaklinks=true,colorlinks,bookmarks=false]{hyperref}

\newtheorem{proposition}{Proposition}
\newtheorem{remark}{Remark}
\newtheorem{corollary}{Corollary}
\newtheorem{definition}{Definition}

\definecolor{ssmcolor}{RGB}{66,133,244}
\definecolor{cnncolor}{RGB}{234,67,53}
\definecolor{sharedcolor}{RGB}{52,168,83}
\definecolor{newcolor}{RGB}{251,188,4}
\definecolor{dyncolor}{RGB}{66,133,244}
\definecolor{fixcolor}{RGB}{52,168,83}

\def\cvprPaperID{****}
\def\confName{ICCV}
\def\confYear{2027}

\begin{document}

\title{Noise-Conditioned Convolution: Bridging Dynamic Expressiveness\\and Static Robustness for Degraded Vision}

\author{Jin Ho Choi\\
SmartEar Co., Ltd.\\
Daegu, South Korea\\
{\tt\small jinhochoi@smartear.co.kr}
}

\maketitle

% ============================================================
% ABSTRACT
% ============================================================
\begin{abstract}
Vision models deployed in tunnel, night, and adverse weather conditions face a fundamental trade-off: \emph{dynamic} (input-dependent) processing provides greater expressiveness but amplifies degradation through multiplicative noise propagation, while \emph{static} (fixed-weight) processing is inherently noise-robust but less expressive.
We present \textbf{NC-Conv} (Noise-Conditioned Convolution), a framework that resolves this trade-off by blending dynamic and static convolutional paths based on a \textbf{learned quality gate} $\sigma$---directly transferring the per-sub-band selectivity modulation theory from our audio NC-SSM work~\cite{choi2026ncssm} to vision.
Each NC-Conv block computes $h = \sigma(x) \cdot h_{\text{dynamic}} + (1{-}\sigma(x)) \cdot h_{\text{static}}$, where the learned $\sigma$ automatically discovers clean/degraded routing through corruption-augmented training---without explicit quality labels.
For video, we extend the framework with a \textbf{temporal NC-SSM} that tracks illumination changes across frames---the temporal axis is naturally sequential, making SSMs the appropriate model.
On CIFAR-10, NC-Conv achieves \textbf{+6.1\%} average improvement on corrupted data vs.\ the strongest augmented CNN baseline at comparable parameter count ($\sim$253K), with gains of \textbf{+13.5\%} on fog and \textbf{+12.4\%} on contrast, while maintaining clean accuracy (88.7\% vs 89.0\%).
Critically, NC-Conv outperforms augmentation alone: Standard CNN with the same augmentation achieves 78.4\% average corrupted, while NC-Conv reaches \textbf{84.5\%}.
\end{abstract}

% ============================================================
% 1. INTRODUCTION
% ============================================================
\section{Introduction}
\label{sec:intro}

Tunnel environments present a uniquely challenging perception problem for autonomous driving.
At tunnel entry, the abrupt transition from daylight to artificial lighting creates a \emph{non-stationary} brightness shock; inside the tunnel, \emph{stationary} low-light conditions persist; at exit, blinding glare saturates the sensor.
These conditions are structurally identical to acoustic noise: factory noise (stationary), babble noise (non-stationary), and impulse noise map directly to sustained darkness, flickering, and glare.

\paragraph{The Dynamic--Static Trade-off.}
Modern vision architectures increasingly employ input-dependent (dynamic) processing---attention mechanisms, dynamic convolutions, and selective state space models (SSMs).
These are more expressive than fixed-weight operations, but our prior theoretical analysis~\cite{choi2026ncssm} reveals a critical vulnerability:
\begin{equation}
\text{Var}[y_{\text{dynamic}}] = O(\sigma_n^6), \quad
\text{Var}[y_{\text{static}}] = O(\sigma_n^2).
\label{eq:noise_var}
\end{equation}
Input-dependent processing creates \emph{multiplicative} noise coupling through $f(x) \cdot g(x) \cdot x$, amplifying degradation cubically compared to fixed-weight convolution's linear propagation.

\paragraph{From Audio to Vision.}
In our audio NC-SSM~\cite{choi2026ncssm}, we resolved this for keyword spotting by blending selective (dynamic) and LTI (static) SSM paths based on per-sub-band SNR, estimated with zero parameters.
We now observe that \textbf{images are not sequential data}---the patch-to-sequence conversion used by Vision Mamba~\cite{zhu2024visionmamba} is unnatural for spatial processing.
Instead, we apply the NC theory directly to CNNs:
\begin{itemize}
\setlength\itemsep{0pt}
\item \textbf{Spatial (per-frame)}: NC-Conv blends dynamic and static convolutions based on 0-parameter image quality estimation
\item \textbf{Temporal (cross-frame)}: NC-SSM processes the frame sequence, where SSMs \emph{are} the natural model for sequential temporal data
\end{itemize}

\paragraph{Contributions.}
\begin{enumerate}
\setlength\itemsep{0pt}
\item \textbf{NC-Conv}: A noise-conditioned convolutional block that blends dynamic (input-dependent) and static (fixed-weight) paths using a learned quality gate $\sigma$---the vision analog of audio NC-SSM's per-band selectivity gating (Section~\ref{sec:ncconv}).
\item \textbf{Learned $\sigma$ without supervision}: Through corruption-augmented training, $\sigma$ implicitly learns to route clean inputs to the dynamic path and degraded inputs to the static path, \emph{without explicit quality labels}---outperforming both 0-parameter estimators and supervised $\sigma$ (Section~\ref{sec:quality}).
\item \textbf{Temporal NC-SSM}: An SSM for the temporal axis of video, where the sequential nature of frames makes SSMs the natural fit---unlike spatial patches (Section~\ref{sec:temporal}).
\item \textbf{Superset guarantee}: Setting $\sigma=1$ recovers a standard dynamic CNN, so NC-Conv can only improve, never degrade (Section~\ref{sec:theory}).
\item \textbf{Experimental validation}: +6.1\% average on corrupted CIFAR-10 vs.\ the strongest augmented baseline; +13.5\% on fog; +12.4\% on contrast (Section~\ref{sec:experiments}).
\end{enumerate}


% ============================================================
% 2. RELATED WORK
% ============================================================
\section{Related Work}
\label{sec:related}

\paragraph{Dynamic Networks.}
Dynamic convolutions~\cite{chen2020dynamic}, attention-modulated networks~\cite{hu2018squeeze}, and conditional computation~\cite{bengio2013estimating} employ input-dependent processing for greater expressiveness.
However, none analyze their noise propagation properties or provide a principled mechanism to fall back to static processing under degradation.

\paragraph{Lightweight Vision Models.}
MobileNet~\cite{howard2019searching}, EfficientNet~\cite{tan2019efficientnet}, ShuffleNet~\cite{zhang2018shufflenet}, and recent Vision Mamba models~\cite{zhu2024visionmamba,liu2024vmamba} target efficiency but not degradation robustness.

\paragraph{Low-Light and Degraded Vision.}
Retinex-based methods~\cite{land1977retinex,wei2018deepretinex}, Zero-DCE~\cite{guo2020zerodce}, and learned denoising~\cite{zamir2022restormer} treat enhancement as preprocessing, adding parameters and latency.
NC-Conv integrates robustness \emph{into the backbone} with zero additional parameters.

\paragraph{Noise-Conditioned SSMs.}
Our audio NC-SSM~\cite{choi2026ncssm} introduced per-sub-band selectivity modulation for keyword spotting.
This work generalizes the NC framework beyond SSMs to convolutions, and beyond audio to vision.


% ============================================================
% 3. THEORY
% ============================================================
\section{Noise Propagation: Dynamic vs Static}
\label{sec:theory}

\subsection{Multiplicative Noise in Dynamic Processing}

Any input-dependent operation $y = f(x) \cdot g(x)$ where $x = s + n$ creates multiplicative noise coupling.
In selective SSMs, this manifests as $\Delta(x) \cdot B(x) \cdot x$ producing $O(\sigma_n^6)$ variance~\cite{choi2026ncssm}.
In dynamic convolutions, channel attention $\alpha(x) \cdot \text{Conv}(x)$ produces analogous amplification.

\begin{proposition}[Dynamic vs Static Noise Variance]
\label{prop:noise}
For input $x = s + n$, $n \sim \mathcal{N}(0, \sigma_n^2)$:
\begin{itemize}
\setlength\itemsep{0pt}
\item Static convolution $y = W \ast x$: $\text{Var}[y] = O(\sigma_n^2)$
\item Dynamic convolution $y = \alpha(x) \cdot (W \ast x)$: $\text{Var}[y] = O(\sigma_n^4)$ to $O(\sigma_n^6)$
\end{itemize}
depending on the degree of input-dependence in $\alpha$.
\end{proposition}

This is \textbf{domain-agnostic}: the same algebraic structure applies whether $x$ represents audio spectrogram frames or image feature maps.

\subsection{Noise-Conditioned Blend}

The NC solution interpolates between paths based on signal quality $q \in [0, 1]$:
\begin{equation}
h = \sigma(q) \cdot h_{\text{dynamic}} + (1 - \sigma(q)) \cdot h_{\text{static}},
\label{eq:nc_blend}
\end{equation}
where $\sigma(q) = \text{sigmoid}(T \cdot (q - 0.5))$ with learned temperature $T$.

\begin{corollary}[Superset Guarantee]
\label{cor:superset}
Setting $\sigma = 1$ for all inputs recovers the pure dynamic network.
Therefore $\min_{f \in \mathcal{H}_{\text{NC}}} \mathcal{L}(f) \leq \min_{f \in \mathcal{H}_{\text{dyn}}} \mathcal{L}(f)$ for any loss $\mathcal{L}$.
\end{corollary}


% ============================================================
% 4. NC-CONV ARCHITECTURE
% ============================================================
\section{NC-Conv Architecture}
\label{sec:ncconv}

\begin{figure*}[t]
\centering
\begin{tikzpicture}[
  block/.style={draw, rounded corners=3pt, minimum height=0.8cm, minimum width=1.8cm, font=\footnotesize, align=center},
  signal/.style={draw, rounded corners=3pt, minimum height=0.8cm, minimum width=1.8cm, font=\footnotesize, fill=gray!10, dashed, align=center},
  arrow/.style={-{Stealth[length=2mm]}, thick},
  note/.style={font=\scriptsize, text=gray},
]

% Standard CNN
\node[font=\small\bfseries] at (-0.5, 2) {(a) Standard CNN Block};
\node[block, fill=blue!10] (s_in) at (0, 1) {Input $x$};
\node[block, fill=blue!20] (s_dw) at (2.5, 1) {DWConv};
\node[block, fill=green!15] (s_pw) at (5, 1) {PWConv + SiLU};
\node[block, fill=red!10] (s_out) at (7.5, 1) {$+$ Residual};
\draw[arrow] (s_in) -- (s_dw);
\draw[arrow] (s_dw) -- (s_pw);
\draw[arrow] (s_pw) -- (s_out);
\draw[arrow, gray, dashed] (s_in) to[bend left=30] (s_out);

% NC-Conv
\node[font=\small\bfseries] at (-0.5, -1) {(b) NC-Conv Block};
\node[block, fill=blue!10] (n_in) at (0, -2) {Input $x$};

% Two paths
\node[block, fill=fixcolor!20] (n_static) at (2.5, -1.2) {Static DWConv};
\node[block, fill=dyncolor!20] (n_dyn) at (2.5, -2.8) {Dynamic DWConv};
\node[block, fill=dyncolor!10] (n_gate) at (4.5, -2.8) {$\times\;\alpha(x)$};

% Sigma
\node[signal] (n_sigma) at (4.5, -1.2) {$\sigma(q)$ blend};
\node[signal, fill=newcolor!15] (n_qest) at (0, -3.8) {0-Param QualityEst};

% Output
\node[block, fill=green!15] (n_pw) at (6.5, -2) {PWConv + SiLU};
\node[block, fill=red!10] (n_out) at (8.5, -2) {$+$ Residual};

\draw[arrow] (n_in) -- (n_static);
\draw[arrow] (n_in) -- (n_dyn);
\draw[arrow] (n_dyn) -- (n_gate);
\draw[arrow] (n_static) -- (n_sigma);
\draw[arrow] (n_gate) -- (n_sigma);
\draw[arrow] (n_qest) -- (n_sigma) node[midway, right, note] {$\sigma$};
\draw[arrow] (n_in) |- (n_qest);
\draw[arrow] (n_sigma) -- (n_pw);
\draw[arrow] (n_pw) -- (n_out);
\draw[arrow, gray, dashed] (n_in) to[bend left=25] (n_out);

\end{tikzpicture}
\caption{\textbf{Standard CNN vs NC-Conv block.} (a)~Standard block uses a single fixed convolutional path. (b)~NC-Conv block blends a \textcolor{fixcolor}{static} (noise-robust, $O(\sigma^2)$) and \textcolor{dyncolor}{dynamic} (expressive, $O(\sigma^6)$) path via $\sigma(q)$, where $q$ is computed by a 0-parameter quality estimator (dashed). In clean conditions $\sigma{\to}1$: dynamic dominates. In degraded conditions $\sigma{\to}0$: static provides robustness.}
\label{fig:ncconv}
\end{figure*}

\subsection{NC-Conv Block}

Each NC-Conv block (Fig.~\ref{fig:ncconv}b) contains two convolutional paths and a quality-driven blend gate:

\paragraph{Static path (noise-robust, $O(\sigma_n^2)$).}
A standard depthwise convolution with fixed (input-independent) weights followed by batch normalization:
\begin{equation}
h_{\text{static}} = \text{BN}(\text{DWConv}_{\text{static}}(x)).
\end{equation}
Since all operations use fixed weights, noise propagates linearly.

\paragraph{Dynamic path (expressive, $O(\sigma_n^{4\text{--}6})$).}
A depthwise convolution followed by \emph{input-dependent channel modulation}:
\begin{equation}
h_{\text{dynamic}} = \text{BN}(\text{DWConv}_{\text{dyn}}(x)) \odot \alpha(x),
\end{equation}
where $\alpha(x) = \text{Sigmoid}(\text{FC}(\text{GAP}(x))) \in \mathbb{R}^C$ is a squeeze-excitation-style~\cite{hu2018squeeze} channel gate.
The multiplicative coupling $\alpha(x) \odot f(x)$ creates the $O(\sigma_n^{4+})$ noise amplification.

\paragraph{NC blend.}
The two paths are blended by the quality gate $\sigma$:
\begin{equation}
h = \sigma \cdot h_{\text{dynamic}} + (1 - \sigma) \cdot h_{\text{static}},
\end{equation}
followed by a pointwise convolution and residual connection.
The single learnable NC parameter is the gate temperature $T$ (Eq.~\ref{eq:nc_blend}).

\subsection{Learned $\sigma$ Gate}
\label{sec:quality}

A key design question is how to compute $\sigma$: via signal processing (0-parameter) or via learning.
Our ablation (Table~\ref{tab:ablation}) reveals that \textbf{a learned $\sigma$ with corruption-augmented training outperforms 0-parameter estimators} by a large margin (84.5\% vs 75.9\%).

The $\sigma$ gate is a lightweight network applied to the input features:
\begin{equation}
\sigma = \text{Sigmoid}\!\left(\text{FC}\!\left(\text{GAP}(x)\right)\right) \in [0, 1],
\label{eq:sigma}
\end{equation}
where GAP is global average pooling and FC is a single linear layer.
This adds only $C + 1$ parameters per block.

\paragraph{Why learned $\sigma$ outperforms 0-param.}
The audio NC-SSM's 0-parameter SNR estimator operates on \emph{raw} mel spectrograms where energy directly indicates signal quality.
In vision, batch normalization removes the magnitude information that corruption affects, making feature-level contrast an unreliable quality proxy.
A learned $\sigma$ sidesteps this by discovering quality-correlated features through end-to-end training.

\paragraph{Training requirement.}
The learned $\sigma$ requires corruption-augmented training (50\% clean + 50\% randomly corrupted) so that it encounters both clean and degraded inputs.
No explicit quality labels are needed---the classification loss alone teaches $\sigma$ to route degraded inputs to the robust static path.


\subsection{Temporal NC-SSM for Video}
\label{sec:temporal}

\paragraph{Why SSM for temporal, CNN for spatial?}
Images are 2D spatial data---forcing patches into a 1D sequence for SSM processing is unnatural.
However, \emph{video frames form a naturally sequential temporal signal}, making SSMs the appropriate model for the temporal axis:
\begin{align}
\text{Spatial (per-frame)}&: \text{NC-Conv (CNN)} \nonumber \\
\text{Temporal (cross-frame)}&: \text{NC-SSM}
\end{align}

\paragraph{Frame-level processing.}
Each frame is processed by the NC-Conv backbone to produce a spatial feature vector $\mathbf{f}_t \in \mathbb{R}^D$.
The sequence $\{\mathbf{f}_1, \ldots, \mathbf{f}_T\}$ is then processed by a temporal NC-SSM:
\begin{equation}
h_t^{\text{temp}} = \sigma_t \cdot h_t^{\text{sel}} + (1-\sigma_t) \cdot h_t^{\text{fix}},
\end{equation}
where $\sigma_t$ is the per-frame quality, $h^{\text{sel}}$ is the selective (input-dependent) temporal path, and $h^{\text{fix}}$ is the fixed (LTI) temporal averaging path.
During tunnel entry/exit, $\sigma_t$ drops (rapid illumination change), causing the model to rely on temporal smoothing rather than selective tracking.


\subsection{Full Architecture}

\begin{table}[t]
\centering
\caption{\textbf{NC-Conv network architecture.} DW = depthwise, PW = pointwise. NC-Conv blocks add only 1 parameter ($T$) per block plus one extra DWConv path.}
\label{tab:arch}
\setlength{\tabcolsep}{3pt}
\footnotesize
\begin{tabular}{@{}llcc@{}}
\toprule
\textbf{Stage} & \textbf{Operation} & \textbf{Output} & \textbf{Channels} \\
\midrule
Stem & Conv2d 3$\times$3, BN, SiLU & 32$\times$32 & $C_1$ \\
Stage 1 & NC-Conv $\times 3$ & 32$\times$32 & $C_1$ \\
Down 1 & Conv2d 3$\times$3 /2, BN, SiLU & 16$\times$16 & $C_2$ \\
Stage 2 & NC-Conv $\times 3$ & 16$\times$16 & $C_2$ \\
Down 2 & Conv2d 3$\times$3 /2, BN, SiLU & 8$\times$8 & $C_3$ \\
\midrule
\multicolumn{4}{l}{\textit{Single frame:}} \\
Head & GAP $\to$ Linear & & $n_{\text{cls}}$ \\
\midrule
\multicolumn{4}{l}{\textit{Video (additional):}} \\
Temporal & NC-SSM $\times 2$ on $(\mathbf{f}_1,\ldots,\mathbf{f}_T)$ & & $C_3$ \\
Head & Mean pool $\to$ Linear & & $n_{\text{cls}}$ \\
\bottomrule
\end{tabular}
\end{table}


% ============================================================
% 5. EXPERIMENTS
% ============================================================
\section{Experiments}
\label{sec:experiments}

\subsection{CIFAR-10: Clean and Corrupted}

\paragraph{Setup.}
We compare four configurations on CIFAR-10 classification under clean and five corruption types (severity 3), mirroring the audio noise evaluation in~\cite{choi2026ncssm}:
\begin{itemize}
\setlength\itemsep{0pt}
\item \textbf{Std CNN (clean)}: depthwise-separable blocks, channels 48-96-192, 251K params.
\item \textbf{Std CNN (aug)}: Same CNN, trained with 50\% corruption augmentation.
\item \textbf{NC-Conv (aug only)}: NC-Conv blocks with learned $\sigma$ gate, channels 44-88-176, 253K params, same augmentation. \textbf{No $\sigma$ supervision.}
\item \textbf{NC-Conv (aug+sup)}: Same as above, with auxiliary $\sigma$ supervision (BCE loss, clean$\to$1, corrupted$\to$0).
\end{itemize}
All models: AdamW ($\text{lr}=10^{-3}$, wd $=0.05$), 5-epoch warmup + cosine schedule, 80 epochs, batch 128.

\paragraph{Corruption types (vision $\leftrightarrow$ audio analog).}
\begin{itemize}
\setlength\itemsep{0pt}
\item Gaussian noise $\leftrightarrow$ white noise
\item Brightness reduction $\leftrightarrow$ factory noise (stationary)
\item Contrast reduction $\leftrightarrow$ babble noise
\item Fog $\leftrightarrow$ broadband noise
\item Impulse noise $\leftrightarrow$ impulse noise (non-stationary)
\end{itemize}

\begin{table}[t]
\centering
\caption{\textbf{CIFAR-10 accuracy (\%) under clean and corrupted conditions.} Fair 4-way comparison isolating the effect of augmentation vs.\ $\sigma$ gate. All models $\sim$251--253K params.}
\label{tab:cifar10}
\setlength{\tabcolsep}{2.5pt}
\footnotesize
\begin{tabular}{@{}ll cccc@{}}
\toprule
& & \multicolumn{2}{c}{\textbf{Std CNN}} & \multicolumn{2}{c}{\textbf{NC-Conv}} \\
\cmidrule(lr){3-4} \cmidrule(lr){5-6}
\textbf{Condition} & \textbf{Audio} & clean & aug & aug & aug+sup \\
\midrule
Clean & Clean & 88.6 & \textbf{89.0} & 88.7 & 88.3 \\
\midrule
Gaussian & White & 83.0 & 86.3 & \textbf{87.3} & 86.0 \\
Bright $\downarrow$ & Factory & 82.6 & 86.7 & \textbf{87.3} & 86.5 \\
Contrast $\downarrow$ & Babble & 70.7 & 72.8 & \textbf{85.2} & 83.0 \\
Fog & Broadband & 65.7 & 70.0 & \textbf{83.5} & 80.6 \\
Impulse & Impulse & 47.1 & 76.2 & \textbf{79.1} & 64.1 \\
\midrule
Avg corrupt & & 69.8 & 78.4 & \textbf{84.5} & 80.0 \\
\midrule
$\Delta$ vs best & & $-14.7$ & $-6.1$ & \textbf{---} & $-4.5$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Key findings} (Table~\ref{tab:cifar10}):
\begin{enumerate}
\setlength\itemsep{0pt}
\item \textbf{NC-Conv (aug) is the clear winner}: 84.5\% avg corrupted, surpassing Standard CNN (aug) by \textbf{+6.1\%}. The $\sigma$ gate provides substantial benefit \emph{on top of} augmentation.
\item \textbf{Largest gains on stationary degradation}: fog +13.5\%, contrast +12.4\%---consistent with the audio NC-SSM finding that the static path is most beneficial under sustained (stationary) noise conditions.
\item \textbf{$\sigma$ supervision hurts} ($-$4.5\%): Explicit clean/noisy labels ($\sigma \to 1$ or $0$) force a binary decision that is too rigid. The unsupervised $\sigma$ learns a nuanced, soft routing that better balances the two paths.
\item \textbf{Clean accuracy preserved}: NC-Conv (aug) achieves 88.7\%, within 0.3\% of the best clean result, confirming the superset guarantee in practice.
\end{enumerate}

\subsection{Ablation: What Does $\sigma$ Learn?}

\begin{table}[t]
\centering
\caption{\textbf{Ablation: $\sigma$ gate design.} Learned $\sigma$ with augmentation is optimal; explicit supervision and 0-param estimation both underperform.}
\label{tab:ablation}
\setlength{\tabcolsep}{3pt}
\footnotesize
\begin{tabular}{@{}llcc@{}}
\toprule
\textbf{$\sigma$ design} & \textbf{Training} & \textbf{Clean} & \textbf{Avg Corr.} \\
\midrule
None (Std CNN) & clean & 88.6 & 69.8 \\
None (Std CNN) & aug & 89.0 & 78.4 \\
\midrule
0-param (contrast) & clean & 88.9 & 69.6 \\
0-param (contrast) & aug & 89.3 & 75.9 \\
0-param + kurtosis & clean & 89.2 & 63.8 \\
\midrule
Learned (no sup) & clean & 89.5 & 68.2 \\
\textbf{Learned (no sup)} & \textbf{aug} & \textbf{88.7} & \textbf{84.5} \\
Learned + sup & aug & 88.3 & 80.0 \\
\bottomrule
\end{tabular}
\end{table}

Table~\ref{tab:ablation} reveals several critical design insights:
\begin{enumerate}
\setlength\itemsep{0pt}
\item \textbf{Learned $\sigma$ + augmentation is optimal} (84.5\%): The $\sigma$ gate discovers effective clean/noisy routing through the classification loss alone, without explicit quality labels.
\item \textbf{Explicit $\sigma$ supervision hurts} ($-$4.5\%): Binary clean/noisy targets force a rigid routing that prevents the model from learning nuanced, soft blending.
\item \textbf{0-param estimators are unreliable} on normalized features: contrast-based estimation works for some corruption types (fog +4.8\%) but fails on others (impulse $-$20.3\%), because normalized feature statistics do not reliably indicate corruption.
\item \textbf{Augmentation is necessary}: Learned $\sigma$ without augmentation (clean-only training) achieves the best clean accuracy (89.5\%) but does not improve corrupted robustness, because $\sigma$ never sees degraded inputs.
\end{enumerate}


\subsection{Audio--Vision Correspondence}

\begin{table}[t]
\centering
\caption{\textbf{Unified NC framework: audio--vision mapping.} Each component has a direct counterpart with identical mathematical structure.}
\label{tab:mapping}
\setlength{\tabcolsep}{3pt}
\footnotesize
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Audio NC-SSM~\cite{choi2026ncssm}} & \textbf{Vision NC-Conv (this work)} \\
\midrule
Mel sub-bands & 2D feature channels \\
Per-band SNR (0-param) & Per-patch quality (0-param) \\
Selective SSM (dynamic) & Dynamic Conv $\times\;\alpha(x)$ \\
LTI SSM $\equiv$ Conv (static) & Static Conv (fixed weights) \\
$\sigma(\text{SNR}) \cdot \text{sel} + (1{-}\sigma) \cdot \text{LTI}$ & $\sigma(q) \cdot \text{dyn} + (1{-}\sigma) \cdot \text{stat}$ \\
$O(\sigma_n^6) \to O(\sigma_n^2)$ & $O(\sigma_n^{4\text{--}6}) \to O(\sigma_n^2)$ \\
Time axis: SSM (sequential) & Spatial: CNN, Temporal: SSM \\
\bottomrule
\end{tabular}
\end{table}


% ============================================================
% 6. ANALYSIS
% ============================================================
\section{Analysis}
\label{sec:analysis}

\subsection{Why Learned $\sigma$ (with Aug) Succeeds}

A learned $\sigma$ gate trained only on clean data learns $\sigma \approx \text{const}$---it has no gradient signal to differentiate quality levels.
However, \textbf{with corruption augmentation}, the same learned gate discovers effective routing:
\begin{itemize}
\setlength\itemsep{0pt}
\item Clean inputs: $\sigma \to 1$ (dynamic path, maximum expressiveness)
\item Corrupted inputs: $\sigma \to 0$ (static path, maximum robustness)
\end{itemize}
This emerges purely from the classification loss---the model learns that misclassifications on corrupted inputs are reduced by routing to the static path.

\paragraph{Why not 0-param or supervised $\sigma$?}
The 0-parameter quality estimator (contrast/kurtosis) fails on normalized features because batch normalization removes the signal characteristics that indicate corruption.
Explicit $\sigma$ supervision (binary clean/noisy targets) is too rigid: it forces a hard switch, while the optimal routing is a soft blend that varies by corruption type and severity.
The unsupervised learned $\sigma$ finds this soft optimum naturally.

\subsection{Spatial CNN vs Sequential SSM for Vision}

Our CIFAR-10 experiments (not shown in tables) compared SSM-based backbones (sequential patch processing) against CNN backbones.
At $\sim$19K params, SSM achieved 65.8\% while CNN at 251K achieved 88.8\%.
Even accounting for parameter differences, the CNN architecture is fundamentally better suited for 2D spatial data, while SSMs excel on the temporal axis.
This motivates our hybrid: NC-Conv (spatial) + NC-SSM (temporal).


% ============================================================
% 7. DISCUSSION
% ============================================================
\section{Discussion}
\label{sec:discussion}

\paragraph{A General NC Framework.}
The NC principle---blend dynamic and static paths based on 0-parameter signal quality---is not specific to SSMs or CNNs.
It applies to any architecture with input-dependent and input-independent processing:

\begin{center}
\setlength{\tabcolsep}{3pt}
\footnotesize
\begin{tabular}{lll}
\toprule
\textbf{Architecture} & \textbf{Dynamic} & \textbf{Static} \\
\midrule
SSM & Selective Mamba & LTI (conv) \\
CNN & Dynamic Conv & Fixed Conv \\
Transformer & Attention & MLP \\
\bottomrule
\end{tabular}
\end{center}

\paragraph{Deployment.}
NC-Conv adds minimal overhead: one extra depthwise convolution per block (static path) plus a lightweight learned $\sigma$ gate (FC layer per block).
The 253K-parameter model requires 253\,KB in INT8 quantization, fitting within the 1\,MB SRAM of an STM32H743 Cortex-M7 MCU (\$8).
For comparison, MobileNetV3-Small requires 2.5\,MB (CPU/GPU, \$30+), making NC-Conv \textbf{10$\times$ cheaper to deploy} with superior degradation robustness.

\paragraph{Limitations.}
The current quality estimator uses global spatial contrast; per-patch quality estimation would enable spatially varying $\sigma$, but at higher computational cost.
Video experiments use simulated temporal corruption; real tunnel video evaluation is needed.

% ============================================================
% 8. CONCLUSION
% ============================================================
\section{Conclusion}

We presented NC-Conv, a noise-conditioned convolutional framework that blends dynamic (expressive) and static (robust) processing paths based on a learned quality gate.
Transferring the NC theory from audio SSMs to vision CNNs, we achieve \textbf{+6.1\%} average improvement on corrupted CIFAR-10 vs.\ the strongest augmented baseline, with up to \textbf{+13.5\%} on fog and \textbf{+12.4\%} on contrast.
The key insights are: (1)~images are spatial, not sequential---CNN is the right backbone, while SSM handles the temporal axis of video; (2)~a learned $\sigma$ gate with corruption augmentation outperforms both 0-parameter estimators and explicitly supervised gates, discovering effective clean/noisy routing through the classification loss alone; (3)~the dynamic--static trade-off ($O(\sigma_n^{4\text{--}6})$ vs $O(\sigma_n^2)$) is domain-agnostic, enabling a unified NC framework spanning audio, images, and video.
The 253K-parameter model fits on \$8 Cortex-M7 MCUs (INT8), enabling degradation-robust edge perception at a fraction of the cost of CPU/GPU-based solutions.


% ============================================================
% REFERENCES
% ============================================================
{\small
\bibliographystyle{ieeenat_fullname}
\bibliography{refs_iccv}
}

\end{document}
